Ship-based lidar evaluation of Southern Ocean low clouds in the storm-resolving general circulation model ICON and the ERA5 and MERRA-2 reanalyses

### 2025-07-21 manuscript_agu_rev1.tex
Small fixes

Because the analyzed ICON simulation was free-running (years 2021–2024, inclusive), weather and climate oscillations [such as the El Niño–Southern Oscillation (ENSO) phase] are not expected to be equivalent to reality. To compare with the observations collected during a different time period (years 2010–2021, inclusive), we compared the model output with observations at the same time of year and geographical location, as determined for each data point, such as a lidar profile or a radiosonde launch. In the ALCF, this was done using the \emph{override\_year} option (\url{https://alcf.peterkuma.net/documentation/cli/cmd_model.html}). For radiosonde profiles, the same mapping of time was done. That is, when selecting an equivalent profile from the model, the time of the profile was changed so that the time relative to the start of the year was preserved, but the year was changed to one of the four years available in the model data. Thus, for every radiosonde launch, there were four equivalent model profiles. The geographical location was kept the same.

@@

ERA5 \cite{era5} is a reanalysis produced by the ECMWF. It is based on an NWP model IFS version CY41R2. It uses the \citeA{tiedtke1993} prognostic cloud scheme and \citeA{forbes2014} for mixed-phase clouds. The horizontal resolution is 0.25° in latitude and longitude, which is about 28~km in the north–south direction and 14~km in the east–west direction at 60°S. Internally, the model uses 137 vertical levels. Here, we use output at 1-hourly instantaneous time intervals, except for radiation quantities, which are accumulations (from these we calculate daily means). Vertically resolved quantities are made available on 37 pressure levels.

@@

Radiation and water path calculations presented in the results (Section~\ref{sec:results}) were completed such that they always represent daily means in order to be consistent with the CERES SYN1deg data. Therefore, every instantaneous profile in the simulated lidar data was assigned a daily mean radiation and water path value corresponding to the day (in the Coordinated Universal Time; UTC). In turn, the average radiation and water paths during the entire voyage or station observation period were calculated as averages of the profile values. In the observed lidar data, the daily mean values were taken from the spatially and temporally co-located CERES SYN1deg data for the day (in UTC). The voyage or station average was calculated in the same way.

@@

\begin{table}[t]
\caption{Summary of the main biases. Values are relative to observations and rounded to the nearest 5, except for daily cloud cover and RH, which are rounded to the nearest integer. The best-performing value is marked in \textbf{bold}. Abbreviations: boundary layer (BL), relative humidity (RH), shortwave (SW), longwave (LW), liquid water path (LWP), ice water path (IWP), and lifting condensation level (LCL).}
\label{tab:summary}
\centering
\begin{tabular}{llll}
& ICON & MERRA-2 & ERA5\\
\hline
Total cloud fraction (\%) & \textbf{-10} & -20 & -20\\
Daily cloud cover (okta) & \textbf{-1} & -2 & -2\\
Fog (\%) & \textbf{0} & -10 & -10\\
BL clouds (at $\sim$500 m) & 15 & \textbf{0} & 5\\
Mid-lev. clouds (at $\sim$1.5 km) & -5 & -5 & \textbf{0}\\
RH at 500 m & 2 & 2 & \textbf{0}\\
SW (W m$^{-2}$) & -25 & \textbf{5} & -20\\
LW (W m$^{-2}$) & 5 & 5 & 5\\
LWP (g m$^{-2}$) & \textbf{10} & 20 & -15\\
IWP (g m$^{-2}$) & -30 & -30 & \textbf{-15}\\
LCL distribution peak (m) & 300 & 300 & 300\\
\hline
\end{tabular}
\end{table}

We analyzed a total of about 2400 days of lidar and 2300 radiosonde observations from 31 campaigns and the Macquarie Island sub-Antarctic station, covering the Atlantic, Australian, and New Zealand sectors of the SO over 10 years. This dataset, together with the use of a ground-based lidar simulator, provided a comprehensive basis for evaluating SO cloud and thermodynamic profile biases in the GSRM ICON and the ERA5 and MERRA-2 reanalyses. Our analysis provides a unique evaluation perspective, complementary to satellite observations for evaluating boundary layer clouds and fog, which are predominant in this region. We did not, however, analyze the cloud phase based on ground-based observations. Cloud phase can have a strong impact on the SW radiative transfer due to larger and therefore less numerous hydrometeors (for the same amount of water) scattering much less SW radiation. Especially, the underestimation of fog or very low-level clouds is very substantial in the reanalyses, and we showed that this relates to cloud formation or persistence at RH between 80 and 100\% in the boundary layer in the observations, while in models RH values less than 100\% are associated with clear bins. We subset the dataset by low and high latitude SO bands, cyclonic activity, and stability in order to identify how these conditions influence the biases. The main identified biases are summarized in Table~\ref{tab:summary} and discussed below.

@@

We show that relative to ERA5, the distribution and strength of cyclonic activity over the SO is well represented in ICON, but it displays lower values of LTS. The latter is also manifested in the radiosonde profile comparison (Fig.~\ref{fig:rs-lts-hist}), showing that the $\theta_v$ profiles in ICON are less stable than in the observations over low-latitude SO. It is also manifested in near-surface air temperature, which is overestimated in the 1–7°C range at the radiosonde launch locations (Fig.~\ref{fig:stats-hist-surf}a). The underestimated LTS is linked to the overestimated cloud peak at 500 m in the lidar cloud occurrence comparison (Fig~\ref{fig:cloud-occurrence}f–g). It might also be interacting with the cloud inhomogeneity factor employed in ICON (Section~\ref{sec:icon}), resulting in lower cloud liquid water used in radiative calculations, hence decreased outgoing SW radiation. Based on the $\theta_v$ profile analysis, clouds at 500 m are predominantly convectively driven, and it is therefore expected that a model bias towards weak stability results in an increased cloud formation at this level. The underestimation of clouds above 1 km in ICON does not have a clear physical reason in our analysis and is likely partially or fully caused by stronger obscuration of the simulated lidar signal by the underlying and overestimated clouds in ICON at around 500 m.

>>>>

Because the analyzed ICON simulation was free-running (years 2021–2024, inclusive), weather and climate oscillations [such as the El Niño–Southern Oscillation (ENSO) phase] are not expected to be equivalent to reality. To compare with the observations collected during a different time period (years 2010–2021, inclusive), we compared the model output with observations at the same time of year and geographical location, as determined for each data point, such as a lidar profile or a radiosonde launch. In the ALCF, this was done using the \emph{override\_year} option.

@@

ERA5 \cite{era5} is a reanalysis produced by the ECMWF. It is based on an NWP model IFS version CY41R2. It uses the \citeA{tiedtke1993} prognostic cloud scheme and the \citeA{forbes2014} scheme for mixed-phase clouds. The horizontal resolution is 0.25° in latitude and longitude, which is about 28~km in the north–south direction and 14~km in the east–west direction at 60°S. Internally, the model uses 137 vertical levels. Here, we use output at 1-hourly instantaneous time intervals, except for radiation quantities, which are accumulations (from these we calculate daily means). Vertically resolved quantities are available on 37 pressure levels.

@@

Radiation and water path calculations presented in the results (Section~\ref{sec:results}) were completed such that they always represent daily means in order to be consistent with the CERES SYN1deg data. Therefore, every instantaneous profile in the simulated lidar data was assigned a daily mean radiation and water path value corresponding to the day (in the Coordinated Universal Time; UTC). In turn, the average radiation and water paths during the entire voyage or station observation period were calculated as averages of the profile values. In the observed lidar data, the daily mean values were taken from the spatially and temporally co-located CERES SYN1deg data for the day (in UTC). The voyage and station averages were calculated in the same way.

@@

\begin{table}[t]
\caption{Summary of the main biases. Values are relative to observations and rounded to the nearest multiple of 5, except for daily cloud cover and RH, which are rounded to the nearest integer. The best-performing value is marked in \textbf{bold}. Abbreviations: boundary layer (BL), relative humidity (RH), shortwave (SW), longwave (LW), liquid water path (LWP), ice water path (IWP), and lifting condensation level (LCL).}
\label{tab:summary}
\centering
\begin{tabular}{llll}
& ICON & MERRA-2 & ERA5\\
\hline
Total cloud fraction (\%) & \textbf{-10} & -20 & -20\\
Daily cloud cover (okta) & \textbf{-1} & -2 & -2\\
Fog (\%) & \textbf{0} & -10 & -10\\
BL clouds (at $\sim$500 m) & 15 & \textbf{0} & 5\\
Mid-lev. clouds (at $\sim$1.5 km) & -5 & -5 & \textbf{0}\\
RH at 500 m & 2 & 2 & \textbf{0}\\
SW (W m$^{-2}$) & -25 & \textbf{5} & -20\\
LW (W m$^{-2}$) & 5 & 5 & 5\\
LWP (g m$^{-2}$) & \textbf{10} & 20 & -15\\
IWP (g m$^{-2}$) & -30 & -30 & \textbf{-15}\\
LCL distribution peak (m) & 300 & 300 & 300\\
\hline
\end{tabular}
\end{table}

We analyzed a total of about 2400 days of lidar and 2300 radiosonde observations from 31 campaigns and the Macquarie Island sub-Antarctic station, covering the Atlantic, Australian, and New Zealand sectors of the SO over 10 years. This dataset, together with the use of a ground-based lidar simulator, provided a comprehensive basis for evaluating SO cloud and thermodynamic profile biases in the GSRM ICON and the ERA5 and MERRA-2 reanalyses. Our analysis provides a unique evaluation perspective, complementary to satellite observations for evaluating boundary layer clouds and fog, which are predominant in this region. We did not, however, analyze the cloud phase based on ground-based observations. Cloud phase can have a strong impact on the SW radiative transfer due to larger and therefore less numerous hydrometeors in cold and mixed-phase clouds (for the same amount of water), scattering much less SW radiation. Especially, the underestimation of fog or very low-level clouds is very substantial in the reanalyses, and we showed that this relates to cloud and fog formation or persistence at RH between 80 and 100\% in the boundary layer in the observations, while in models RH values less than 100\% are associated with clear bins. We subset the dataset by low and high latitude SO bands, cyclonic activity, and stability in order to identify how these conditions influence the biases. The main identified biases are summarized in Table~\ref{tab:summary} and discussed below.

@@

We show that relative to ERA5, the distribution and strength of cyclonic activity over the SO is well represented in ICON, but it displays lower values of LTS. The latter is also manifested in the radiosonde profile comparison (Fig.~\ref{fig:rs-lts-hist}), showing that the $\theta_v$ profiles in ICON are less stable than in the observations. It is also manifested in near-surface air temperature, which is overestimated in the 1–7°C range at the radiosonde launch locations (Fig.~\ref{fig:stats-hist-surf}a). The underestimated LTS is linked to the overestimated cloud peak at 500 m in the lidar cloud occurrence comparison (Fig~\ref{fig:cloud-occurrence}f–g). It might also be interacting with the cloud inhomogeneity factor employed in ICON (Section~\ref{sec:icon}), resulting in lower cloud liquid water used in radiative calculations, hence decreased outgoing SW radiation. Based on the $\theta_v$ profile analysis, clouds at 500 m are predominantly convectively driven, and it is therefore expected that a model bias towards weak stability results in an increased cloud formation at this level. The underestimation of clouds above 1 km in ICON does not have a clear physical reason in our analysis and is likely partially or fully caused by stronger obscuration of the simulated lidar signal by the underlying and overestimated clouds in ICON at around 500 m.
